---
title: "Day 1 Notebook"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)

```

```{r}
mydata <- read_lines("Day1Sample1.txt")
mydata
```

## Part 1

According to the instructions, the goal is to take the first number from the left and the first number from the right (this might be the same number) - mash them together and get a list of numbers. So for this - we'll remove all of the letters from the data:

```{r}
part1 <- sapply(mydata,function(x){
  str_replace_all(x,"[abcdefghijklmnopqrstuvwxyz]","")})

unname(part1)
```

grab the first and last numeral and turn them into a single number

```{r}
part1 <- sapply(part1,function(x){
  as.numeric(paste0(str_sub(x,1,1),str_sub(x,-1,-1)))})

unname(part1)
```

then add all of the numbers together to get the final answer

```{r}
part1 <- sum(part1)
part1
```

## Part 2

```{r,include=FALSE}
mydata <- read_lines("Day1Sample2.txt")
```

New sample data:

```{r}
mydata
```

Looking through the data, there are whole word numbers in there. And so now, redo the thing with words to be transformed into numbers. In other words - instead of:

two1nine -\> 1 -\> 11\

the transform is:\
\
two1nine -\> 219 -\> 29

But there's a small hitch - look at the example: \

eigh**t**wothree\

that fifth letter "t" is shared for both the "eight" and "two" and both need to be considered. It is not 8wo3 or eigh23, but 823 - the way you probably read it if you skimmed quickly.

Luckily, from 0-9 (or "zero" to "nine"), the only overlaps are in the first and last letters of the word (Probably for all numbers). So a quick find/replace that saves the first and last letters (e.g.,"three" to "t3e") should catch all of the overlaps

Transform first:

```{r}
part2 <- sapply(mydata,function(x){
  str_replace_all(x,c("one"="o1e",
                      "two"="t2o",
                      "three"="t3e",
                      "four"="f4r",
                      "five"="f5e",
                      "six"="s6x",
                      "seven"="s7n",
                      "eight"="e8t",
                      "nine"="n9e",
                      "zero"="z0o"))})

unname(part2)
```

then follow the remainder of the steps from part 1 to get the answer to part 2

```{r}
unname(part2)

part2 <- sapply(part2,function(x){
  str_replace_all(x,"[abcdefghijklmnopqrstuvwxyz]","")})

unname(part2)

part2 <- sapply(part2,function(x){
  as.numeric(paste0(str_sub(x,1,1),str_sub(x,-1,-1)))})

unname(part2)

part2 <- sum(part2)
part2
```

As a function:

```{r}
day1answers<-function(input){
  part1 <- sapply(input,function(x){
    str_replace_all(x,"[abcdefghijklmnopqrstuvwxyz]","")})
  part1 <- sapply(part1,function(x){
    as.numeric(paste0(str_sub(x,1,1),str_sub(x,-1,-1)))})

  part1 <- sum(part1,na.rm=TRUE)
  
  part2 <- sapply(input,function(x){
    str_replace_all(x,c("one"="o1e",
                        "two"="t2o",
                        "three"="t3e",
                        "four"="f4r",
                        "five"="f5e",
                        "six"="s6x",
                        "seven"="s7n",
                        "eight"="e8t",
                        "nine"="n9e",
                        "zero"="z0o",
                        "[abcdefghijklmnopqrstuvwxyz]"=""))})
  part2 <- sapply(part2,function(x){
    as.numeric(paste0(str_sub(x,1,1),str_sub(x,-1,-1)))})
  part2 <- sum(part2,na.rm=TRUE)
  return(c(part1,part2))}
```

```{r}
day1answers(mydata)
```
