---
title: "Day 9"
output: html_notebook
---
```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(ggplot2)
library(readr)
options(scipen = 999)
```

## Part 1

The goal is to find the next value in each sequence then add them together - 
First, create a tibble, column 1 is a vector of the history values 

```{r}
### Read in histories
input <- read_lines("Day9Sample.txt")

histories<-c()
for(i in 1:length(input)){
  oasis<-as.numeric(unlist(str_split(input[i]," ")))
  histories <- append(histories,list(oasis))}

histories <- tibble(histories,0)
colnames(histories)<-c("history","extrapolate")
```

At each level, the value of extrapolation equals the value to its left, plus the previous value of extrapolation and eventually the values of extrapolation = 0, all you really need to do is add all the values to the left. (or all of the final values of the input vector)
So first, a function that takes in a numeric vector and outputs a vector of the differences between the numbers:

```{r}
onedown<-function(history){
  nextline<-c()
  for(i in 2:length(history)){
    nextline<-c(nextline,(history[i]-history[i-1]))}
  nextline}

onedown(c(0,3,6,9,12,15))
#onedown(unlist(histories[1,1][[1]]))
```
With that function, first apply it over and over until there's only one value left in the line.

```{r}
for (i in 1:(length(oasis)-1)){
histories <- histories %>% rowwise %>%
  mutate(extrapolate=tail(history[[length(history)]][[1]])+extrapolate) %>%
  mutate(history=list(onedown(unlist(history))))}
```
and then, just in case there's a final value, add it to extrapolate (again, it will just add 0 if the differences have already reduced out)

```{r}
histories <- histories %>% rowwise %>%
  mutate(extrapolate=extrapolate+history[[1]])
```
finally, add the extrapolations to get the answer to part 1

```{r}
part1 <- sum(histories$extrapolate)
part1
```
## Part 2

Now, to extrapolate the other side - 
For this,  equals the value to its left, *minus* the previous value of extrapolation - so working its way down.  So it will be add, subtract, add, subtract, etc... down to 0.


```{r, include=FALSE}
histories<-c()
for(i in 1:length(input)){
  oasis<-as.numeric(unlist(str_split(input[i]," ")))
  histories <- append(histories,list(oasis))}

histories <- tibble(histories,0)
colnames(histories)<-c("history","previous")
```

Still use onedown(), but this time, change the previous value

```{r}
for (i in 2:length(oasis)){
histories <- histories %>% rowwise %>%
  mutate(previous=(history[[1]][[1]])*((-1)^i)+previous) %>%
  mutate(history=list(onedown(unlist(history))))}
```
then either add or subtract the last number (just in case it isn't 0)
```{r}
i<-i+1
histories <- histories %>% rowwise %>%
  mutate(previous=(history[[1]]*((-1)^i)+previous))
```
And finally part2 is the sum of previous
```{r}
part2 <- sum(histories$previous)
part2
```
Single function (which needs onedown())

```{r}
day9answers <- function(input){
histories<-c()
for(i in 1:length(input)){
  oasis<-as.numeric(unlist(str_split(input[i]," ")))
  histories <- append(histories,list(oasis))}

histories <- tibble(histories,0)
histories <- cbind(histories,0)
colnames(histories)<-c("history","extrapolate","previous")

for (i in 2:length(oasis)){
histories <- histories %>% rowwise %>%
  mutate(extrapolate=tail(history[[length(history)]][[1]])+extrapolate,
         previous=(history[[1]][[1]])*((-1)^i)+previous) %>%
  mutate(history=list(onedown(unlist(history))))}

histories <- histories %>% rowwise %>%
  mutate(extrapolate=extrapolate+history[[1]]) %>%
  mutate(previous=(history[[1]]*((-1)^i)+previous))

part1 <- sum(histories$extrapolate)
part2 <- sum(histories$previous)
c(part1,part2)}
```


```{r}
day9<- day9answers(input)
day9
```