---
title: "Day 9"
output: html_notebook
---
```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(ggplot2)
library(readr)
options(scipen = 999)
```

## Part 1

The goal is to find the next value in each sequence then add them together - 
First, create a list of vectors (each vector is a history list)

```{r}
### Read in histories
input <- read_lines("Day9Sample.txt")

histories<-c()
for(i in 1:length(input)){oasis<-as.numeric(unlist(str_split(input[i]," ")))
  histories <- append(histories,list(oasis))}
historylength <- length(oasis)
```

At each level, the value of extrapolation equals the value to its left, plus the previous value of extrapolation and eventually the values of extrapolation = 0, all you really need to do is add all the values to the left. (or all of the final values of the input vector)
So first, a function that takes in a numeric vector and outputs a vector of the differences between the numbers:

```{r}
onedown<-function(history){
  nextline<-c()
  for(i in 2:length(history)){
    nextline<-c(nextline,(history[i]-history[i-1]))}
  nextline}

onedown(c(0,3,6,9,12,15))
#onedown(unlist(histories[1,1][[1]]))
```
Add the number on the right and then apply the function over and over until there are no values left in the line.

```{r}

part1<-0
j <- historylength
for (i in 1:historylength){
  lastnums <- sapply(histories,function(x){x[j]})
  part1 <- part1+sum(lastnums)
  histories <- lapply(histories,onedown)
  j<-j-1}

part1
```

## Part 2

Now, to extrapolate the other side - 
For this,  equals the value to its left, *minus* the previous value of extrapolation - so working its way down.  So it will be add, subtract, add, subtract, etc... down to 0.


```{r, include=FALSE}
### this resets histories after running part 1

histories<-c()
for(i in 1:length(input)){
  oasis<-as.numeric(unlist(str_split(input[i]," ")))
  histories <- append(histories,list(oasis))}
```

This time, we go back and forth between adding and subtracting the number on the left

```{r}

part2<-0
for (i in 1:historylength){
  firstnums <- sapply(histories,function(x){x[1]})
  if(i %% 2==1){part2 <- part2+sum(firstnums)
  }else{part2 <- part2-sum(firstnums)}
  histories <- lapply(histories,onedown)}

part2
```
Single function (which needs onedown())

```{r}
day9answers <- function(input){
  histories<-c()
  for(i in 1:length(input)){
    oasis<-as.numeric(unlist(str_split(input[i]," ")))
    histories <- append(histories,list(oasis))}
  
  part1<-0
  part2<-0
  j <- historylength
  for (i in 1:historylength){
    ## get the last numbers, then first numbers
    lastnums <- sapply(histories,function(x){x[j]})
    firstnums <- sapply(histories,function(x){x[1]})
    ## add (or subtract) the sums of the last and firsts to our answers
    part1 <- part1+sum(lastnums)
    if(i %% 2==1){part2 <- part2+sum(firstnums)
    }else{part2 <- part2-sum(firstnums)}
    ##shorten the lists
    histories <- lapply(histories,onedown)
    ## decrement j to find the last number next time.
    j<-j-1}
  c(part1,part2)}
```


```{r}
day9 <- day9answers(input)
day9
```