---
title: "Day 13 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(igraph)
library(ggplot2)
library(rgl)
library(ggraph)
library(readr)
library(ds4psy)
library(numbers)
options(scipen = 999)
```


## Part 1
Find the points of reflection for each frame

Start by setting up the frames:
```{r}
input <- read_lines("Day13Sample.txt")
blanklines <- c(which(input==""))
starts<-c(1,blanklines+1)
ends<-c(blanklines-1,length(input))
frames<-list()

for(i in 1:length(starts)){
    dfm <- matrix(nrow=0,ncol=nchar(input[starts[i]]))
    for(j in starts[i]:ends[i]){
      dfm <- rbind(dfm,c(unlist(str_split(input[j],""))))}
    dfm<-data.frame(dfm)
    frames<-append(frames,list(dfm))}
```

Now to create two functions to find a point of reflection

```{r}
reflecth<-function(x){
  horlist<-c()
  for (i in 1:(nrow(x)-1)){
    ### for each row, see if it is the same as the next row
    if(all(x[i,]==x[(i+1),])){
      j<-i-1
      k<-i+2
      additin<-TRUE
      ### if so, start checking the rows above and below to see if they're also mirrors - until you reach an edge.
      while(j>=1 && k<=(nrow(x))&&additin){
        if((any(x[j,]!=x[k,]))){additin<-FALSE}
        j<-j-1
        k<-k+1}
      ###if all they are all mirrors, add to the list of horizontal reflections
      if (additin){horlist <- c(horlist,i)}}}
  if(is.null(horlist)){horlist<-c(0)}
  horlist}

reflectv<-function(x){
  vertlist<-c()
  for (i in 1:(ncol(x)-1)){
    ### for each column, see if it reflects the column next to it.
    if(all((x[,i]==x[,(i+1)]))){
      ### if so, check columns out to the right & left until you hit an edge.
      j<-i-1
      k<-i+2
      additin<-TRUE
      while(j>=1 && k<=(ncol(x))&&additin){
        if(!(all(x[,j]==x[,k]))){additin<-FALSE}
        j<-j-1
        k<-k+1}
      if(additin==TRUE){vertlist<-c(vertlist,i)}}}
  if(is.null(vertlist)){vertlist<-c(0)}
  vertlist}
```
Find the lists (it is possible there is more than one reflection for horizontal or vertical or both)
```{r}
horizontals <- sapply(frames,reflecth)
verticals <- sapply(frames,reflectv)

part1<-sum(100*horizontals+verticals)
part1
```
## Part 2

It's not the elves fault this time - there's an error, so the goal is to find things that are *almost* mirror images - at most one error.

This means a slight change to the functions:

```{r}
### This is a little sloppy. It doesn't ensure that the error in h & v are the same. And it worked, so I'm letting it go.  

smudgeh<-function(x){
  horlist<-c()
  for (i in 1:(nrow(x)-1)){
    ##smudgecount is the number of differences between lines.  if it is 0 or 1, keep going.
    smudgecount<-sum(x[i,]!=x[(i+1),])
    if(smudgecount<=1){
      j<-i-1
      k<-i+2
      additin<-TRUE
      while(j>=1 && k<=(nrow(x))&&additin){
        ##add the number of differences between the next two lines. if it is still at 0 or 1, keep going.
        smudgecount<-smudgecount+sum(x[j,]!=x[k,])
        if(smudgecount>1){
          additin<-FALSE}
        j<-j-1
        k<-k+1}
      ## only add if exactly 1 smudge
      if (additin&&smudgecount==1){horlist <- c(horlist,i)}}}
  if(is.null(horlist)){horlist <- c(0)}
  horlist}


smudgev<-function(x){
  vertlist<-c()
  for (i in 1:(ncol(x)-1)){
    smudgecount<-sum(x[,i]!=x[,(i+1)])
    if(smudgecount<=1){
      j<-i-1
      k<-i+2
      additin<-TRUE
      while(j>=1 && k<=(ncol(x))&&additin){
        smudgecount<-smudgecount+sum(x[,j]!=x[,k])
        if(smudgecount>1){
          additin<-FALSE}
        j<-j-1
        k<-k+1}
      ## only add if exactly 1 smudge
      if(additin&&smudgecount==1){vertlist<-c(vertlist,i)}}}
  
  if(is.null(vertlist)){vertlist <- c(0)}
  vertlist}
```

and again run and add them together for part 2

```{r}
horiz_smudge <- sapply(frames,smudgeh)
vert_smudge <- sapply(frames,smudgev)

part2<-sum(100*horiz_smudge+vert_smudge)
part2
```

And the single function (it uses a variant of the smudge functions)


```{r}
###helpful functions

comboh<-function(x){
  horlist<-c()
  smudgelist<-c()
  for (i in 1:(nrow(x)-1)){
    ##smudgecount is the number of differences between lines.  if it is 0 or 1, keep going.
    smudgecount<-sum(x[i,]!=x[(i+1),])
    if(smudgecount<=1){
      j<-i-1
      k<-i+2
      additin<-TRUE
      while(j>=1 && k<=(nrow(x))&&additin){
        ##add the number of differences between the next two lines. if it is still at 0 or 1, keep going.
        smudgecount<-smudgecount+sum(x[j,]!=x[k,])
        if(smudgecount>1){
          additin<-FALSE}
        j<-j-1
        k<-k+1}
      ## only add if exactly 1 smudge
      if (additin&&smudgecount==0){horlist <- c(horlist,i)}
      if (additin&&smudgecount==1){smudgelist <- c(smudgelist,i)}}}
  if(is.null(horlist)){horlist <- c(0)}
  if(is.null(smudgelist)){smudgelist <- c(0)}
  append(horlist,smudgelist)}


combov<-function(x){
  vertlist<-c()
  smudgelist<-c()
  for (i in 1:(ncol(x)-1)){
    smudgecount<-sum(x[,i]!=x[,(i+1)])
    if(smudgecount<=1){
      j<-i-1
      k<-i+2
      additin<-TRUE
      while(j>=1 && k<=(ncol(x))&&additin){
        smudgecount<-smudgecount+sum(x[,j]!=x[,k])
        if(smudgecount>1){
          additin<-FALSE}
        j<-j-1
        k<-k+1}
      ## only add if exactly 1 smudge
      if(additin&&smudgecount==0){vertlist<-c(vertlist,i)}
      if(additin&&smudgecount==1){smudgelist<-c(smudgelist,i)}}}
  if(is.null(smudgelist)){smudgelist <- c(0)}
  if(is.null(vertlist)){vertlist <- c(0)}
  append(vertlist,smudgelist)}

### main function

day13answers <-function(input){
  ### setup  
  
  blanklines <- c(which(input==""))
  starts<-c(1,blanklines+1)
  ends<-c(blanklines-1,length(input))
  frames<-list()
  
  for(i in 1:length(starts)){
    dfm <- matrix(nrow=0,ncol=nchar(input[starts[i]]))
    for(j in starts[i]:ends[i]){
      dfm <- rbind(dfm,c(unlist(str_split(input[j],""))))}
    dfm<-data.frame(dfm)
    frames<-append(frames,list(dfm))}

  ### get the hs & vs
  hs<-lapply(frames,comboh)
  vs<-lapply(frames,combov)

  p1h<-sum(sapply(hs,function(x){x[1]}))
  p2h<-sum(sapply(hs,function(x){x[2]}))
  p1v<-sum(sapply(vs,function(x){x[1]}))
  p2v<-sum(sapply(vs,function(x){x[2]}))
    
  part1 <-100*p1h+p1v
  part2 <-100*p2h+p2v
  c(part1,part2)}

```

```{r}
day13 <-day13answers(input)
day13
```